---
title: "Exploratory Data Analysis"
author: "Owais"
date: "2025-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Exploratory Data Analysis

This report performs exploratory data analysis on the `diamonds` and `faithful` datasets from the `ggplot2` package. We'll examine distributions of key variables, identify patterns, and handle outliers.

## Loading Required Packages

```{r packages}
library(tidyverse)
```

## 1. Diamonds Dataset Exploration

### Initial Data Inspection

```{r glimpse}
glimpse(diamonds)
```

The dataset contains `r nrow(diamonds)` observations with `r ncol(diamonds)` variables, including both categorical and continuous variables.

### Categorical Variable Analysis: Cut Quality

#### Bar Plot of Cut Categories

```{r cut_barplot}
ggplot(diamonds) + 
  geom_bar(mapping = aes(x = cut)) +
  labs(title = "Distribution of Diamond Cut Quality",
       x = "Cut Quality",
       y = "Count") +
  theme_minimal()
```

#### Frequency Table of Cut Categories

```{r cut_count}
diamonds %>% 
  count(cut) %>% 
  arrange(desc(n)) %>% 
  knitr::kable(caption = "Frequency of Diamond Cut Categories")
```

### Continuous Variable Analysis: Carat Weight

#### Histogram with 0.5 Carat Bins

```{r carat_hist1}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5) +
  labs(title = "Diamond Carat Weight Distribution",
       subtitle = "Bin width = 0.5 carats",
       x = "Carat Weight",
       y = "Count") +
  theme_minimal()
```

#### Detailed Histogram with 0.1 Carat Bins

```{r carat_hist2}
ggplot(diamonds) + 
  geom_histogram(aes(x = carat), binwidth = 0.1) +
  labs(title = "Detailed Diamond Carat Weight Distribution",
       subtitle = "Bin width = 0.1 carats",
       x = "Carat Weight",
       y = "Count") +
  theme_minimal()
```

#### Broad Overview with 1.0 Carat Bins

```{r carat_hist3}
ggplot(diamonds) + 
  geom_histogram(aes(x = carat), binwidth = 1.0) +
  labs(title = "Broad Diamond Carat Weight Distribution",
       subtitle = "Bin width = 1.0 carats",
       x = "Carat Weight",
       y = "Count") +
  theme_minimal()
```

#### Binned Frequency Count

```{r carat_binned}
diamonds %>% 
  count(cut_width(carat, 0.5)) %>% 
  knitr::kable(caption = "Frequency Count by 0.5 Carat Bins")
```

### Analyzing a Subset Without Outliers

Create a subset with diamonds under 3 carats to remove outliers:

```{r small_subset}
small <- diamonds %>% filter(carat < 3)
```

#### Detailed Histogram of Subset

```{r small_hist}
ggplot(data = small, mapping = aes(x = carat)) + 
  geom_histogram(binwidth = 0.1) +
  labs(title = "Carat Distribution (Diamonds < 3 carats)",
       x = "Carat Weight",
       y = "Count") +
  theme_minimal()
```

#### Frequency Polygon by Cut Quality

```{r freqpoly}
ggplot(data = small, mapping = aes(x = carat, colour = cut)) + 
  geom_freqpoly(binwidth = 0.1) +
  labs(title = "Carat Distribution by Cut Quality",
       x = "Carat Weight",
       y = "Count") +
  theme_minimal()
```

#### High-Resolution Histogram (0.01 carat bins)

```{r tiny_bins}
ggplot(data = small, mapping = aes(x = carat)) + 
  geom_histogram(binwidth = 0.01) +
  labs(title = "High-Resolution Carat Distribution",
       subtitle = "Bin width = 0.01 carats",
       x = "Carat Weight",
       y = "Count") +
  theme_minimal()
```

### Analyzing the 'y' Variable (Width in mm)

#### Initial Histogram

```{r y_hist1}
ggplot(data = diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  labs(title = "Distribution of Diamond Width (y)",
       x = "Width (mm)",
       y = "Count") +
  theme_minimal()
```

#### Zoomed-In Histogram

```{r y_hist2}
ggplot(data = diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50)) +
  labs(title = "Zoomed Distribution of Diamond Width (y)",
       subtitle = "Y-axis limited to 50 counts",
       x = "Width (mm)",
       y = "Count") +
  theme_minimal()
```

#### Identifying Outliers

```{r outliers}
outliers <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>% 
  arrange(y)

knitr::kable(outliers, caption = "Outliers in Diamond Width (y)")
```

## 2. Faithful Dataset Exploration

### Initial Data Inspection

```{r faithful_glimpse}
glimpse(faithful)
```

### Eruption Duration Analysis

```{r faithful_hist}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25) +
  labs(title = "Old Faithful Eruption Duration Distribution",
       x = "Eruption Duration (minutes)",
       y = "Count") +
  theme_minimal()
```















## Conclusion

This exploratory analysis revealed:
- The distribution of diamond characteristics
- Patterns in carat weights
- Outliers in diamond dimensions
- Bimodal distribution in Old Faithful eruption times

Further analysis could investigate relationships between variables (e.g., carat vs. price).